---
title: "ExploratoryVis - Census"
author: "Kyle Szela"
date: "April 24, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
setwd("~/Downloads/census")
census <- read.csv("census-income.csv", stringsAsFactors = T, strip.white = T)

#By Sector
ratio_class_of_worker = census[!(census$class.of.worker == "Not in universe" | 
                                   census$class.of.worker == "Without pay"),]
ratio_class_of_worker = ratio_class_of_worker[(ratio_class_of_worker$full.or.part.time.employment.stat == "Full-time schedules"),]

ratio_class_of_worker = group_by(ratio_class_of_worker, class.of.worker) %>% 
  summarise(., 
            count_geq50 = sum(instance.weight[X == "50000+." & age >= 18]), 
            count_l50 = sum(instance.weight[X == "-50000" & age >= 18]),
            pct_geq50 = (count_geq50 / (count_l50 + count_geq50)) * 100)
classPlot = ggplot(ratio_class_of_worker, aes(x = class.of.worker, y = pct_geq50)) + 
  geom_bar(stat = "identity", aes(fill = class.of.worker)) + 
  scale_fill_brewer(name = "Class of Worker", palette = "Set1") +
  ggtitle("Income by Sector 94' - 95'") + 
  xlab("") + 
  ylab("Percent of Sector with Full-time Income >50k") + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())

#Graph showing income by sector and sex.
ratio_class_of_worker = census[!(census$class.of.worker == "Not in universe" | 
                                   census$class.of.worker == "Without pay"),]
ratio_class_of_worker = ratio_class_of_worker[(ratio_class_of_worker$full.or.part.time.employment.stat == "Full-time schedules"),]

ratio_class_of_worker = group_by(ratio_class_of_worker, class.of.worker, sex) %>% 
  summarise(., 
            count_geq50 = sum(instance.weight[X == "50000+." & age >= 18]), 
            count_l50 = sum(instance.weight[X == "-50000" & age >= 18]),
            pct_geq50 = (count_geq50 / (count_l50 + count_geq50)) * 100)
classPlot_sex = ggplot(ratio_class_of_worker, aes(x = class.of.worker, y = pct_geq50)) + 
  geom_bar(stat = "identity", aes(fill = class.of.worker)) + 
  scale_fill_brewer(name = "Class of Worker", palette = "Set1") + 
  #theme_bw() + 
  ggtitle("Income by Sector 94' - 95'") + 
  xlab("") + 
  ylab("Percent of Sector with Full-time Income >50k") + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  facet_wrap( ~ sex)

#Graph showing income by race.
ratio_race_of_worker = census[(census$full.or.part.time.employment.stat == "Full-time schedules"),]

ratio_race_of_worker = group_by(ratio_race_of_worker, race) %>% 
  summarise(., 
          count_geq50 = sum(instance.weight[X == "50000+." & age >= 18]), 
          count_l50 = sum(instance.weight[X == "-50000" & age >= 18]),
          pct_geq50 = (count_geq50 / (count_l50 + count_geq50)) * 100)
racePlot = ggplot(ratio_race_of_worker, aes(x = race, y = pct_geq50)) + 
  geom_bar(stat = "identity", aes(fill = race)) + 
  scale_fill_brewer(name = "Race", palette = "Set1") +
  ggtitle("Income by Race 94' - 95'") + 
  xlab("") + 
  ylab("Percent of Race with Full-time Income >50k") + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
racePlot

#Graph showing income by race and sex.
ratio_race_of_worker = census[(census$full.or.part.time.employment.stat == "Full-time schedules"),]

ratio_race_of_worker = group_by(ratio_race_of_worker, race, sex) %>% 
  summarise(., 
            count_geq50 = sum(instance.weight[X == "50000+." & age >= 18]), 
            count_l50 = sum(instance.weight[X == "-50000" & age >= 18]),
            pct_geq50 = (count_geq50 / (count_l50 + count_geq50)) * 100)
racePlot_sex = ggplot(ratio_race_of_worker, aes(x = race, y = pct_geq50)) + 
  geom_bar(stat = "identity", aes(fill = race)) + 
  scale_fill_brewer(name = "Race", palette = "Set1") +
  ggtitle("Income by Race 94' - 95'") + 
  xlab("") + 
  ylab("Percent of Race with Full-time Income >50k") + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  facet_wrap( ~ sex)
racePlot_sex

#Graph showing income by age
ratio_age_census = census[(census$full.or.part.time.employment.stat == "Full-time schedules"),]

#Both Sexes
ratio_age = group_by(ratio_age_census, age) %>% 
  summarise(.,
            Both = (sum(instance.weight[X == "50000+."]) / 
                               (sum(instance.weight[X == "50000+."]) + 
                                  sum(instance.weight[X == "-50000"]))) * 100)
#Just males
ratio_age_male = group_by(ratio_age_census, age) %>%
  summarise(.,
            Males = (sum(instance.weight[X == "50000+." & sex == "Male"]) / 
                       (sum(instance.weight[X == "50000+." & sex == "Male"]) + 
                          sum(instance.weight[X == "-50000" & sex == "Male"]))) * 100)
#Just females
ratio_age_female = group_by(ratio_age_census, age) %>%
  summarise(.,
            Females = (sum(instance.weight[X == "50000+." & sex == "Female"]) / 
                       (sum(instance.weight[X == "50000+." & sex == "Female"]) + 
                          sum(instance.weight[X == "-50000" & sex == "Female"]))) * 100)
ratio_age = inner_join(ratio_age, ratio_age_male, by = "age")
ratio_age = inner_join(ratio_age, ratio_age_female, by = "age")
age_Plot = ggplot(ratio_age, aes(age)) +
  stat_smooth(aes(y = Both, colour = "Both"), # continuous x-axis
              se = F, method = "lm", formula = y ~ poly(x, 12)) + 
  stat_smooth(aes(y = Males, colour = "Males"), # continuous x-axis
              se = F, method = "lm", formula = y ~ poly(x, 12)) + 
  stat_smooth(aes(y = Females, colour = "Females"), # continuous x-axis
              se = F, method = "lm", formula = y ~ poly(x, 9)) + 
  theme(legend.title=element_blank()) +
  ggtitle("Income by Age 94' - 95'") + 
  xlab("Age (years)") + 
  ylab("Percent at Age with Full-time Income >50k")

#Looking at female to male education factor and earnings
#First graph number of females and males for each education level
education_numbers = census[(census$full.or.part.time.employment.stat == "Full-time schedules"),]

new_levels = c("Children", "Less than 1st grade", "1st 2nd 3rd or 4th grade", 
               "5th or 6th grade", "7th and 8th grade", "9th grade", "10th grade", 
               "11th grade", "12th grade no diploma", "High school graduate", 
               "Associates degree-academic program", "Associates degree-occup /vocational", 
               "Some college but no degree", "Bachelors degree(BA AB BS)", 
               "Masters degree(MA MS MEng MEd MSW MBA)", "Doctorate degree(PhD EdD)", 
               "Prof school degree (MD DDS DVM LLB JD)")

education_numbers$education = factor(education_numbers$education, levels = new_levels)

education_totals = summarise(education_numbers, 
                             Both = sum(instance.weight),
                             Males = sum(instance.weight[sex == "Male"]),
                             Females = sum(instance.weight[sex == "Female"]))

education_numbers_both = group_by(education_numbers, education) %>%
  summarise(.,
            Both = (sum(instance.weight) / education_totals$Both) * 100)

education_numbers_males = group_by(education_numbers, education, sex) %>%
  summarise(.,
            pct = (sum(instance.weight[sex == "Male"]) / education_totals$Males) * 100)

education_numbers_females = group_by(education_numbers, education, sex) %>%
  summarise(.,
            pct = (sum(instance.weight[sex == "Female"]) / education_totals$Females) * 100)

education_numbers_males$pct = education_numbers_males$pct + education_numbers_females$pct

#Plot for both
educationPlot_both = ggplot(education_numbers_both, aes(x = education, y = Both)) + 
  geom_bar(stat = "identity", aes(fill = education)) + 
  #scale_fill_brewer(name = "Education", palette = "Set1") + 
  theme_bw() + 
  scale_fill_discrete(name = "Education") + 
  ggtitle("Education of Population 94' - 95'") + 
  xlab("") + 
  ylab("Percent of Population with Education") + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
educationPlot_both

#Male/female side by side
educationPlot_mf = ggplot(education_numbers_males, aes(x = education, y = pct)) + 
  geom_bar(stat = "identity", aes(fill = education)) + 
  #scale_fill_brewer(name = "Education", palette = "Set1") + 
  theme_bw() + 
  scale_fill_discrete(name = "Education") + 
  ggtitle("Education of Population 94' - 95'") + 
  xlab("") + 
  ylab("Percent of Population with Education") + 
  theme(axis.ticks = element_blank(), axis.text.x = element_blank()) + 
  facet_wrap( ~ sex)
```

## Data Set

Using the resource, UCI Machine Learning Repository, found the Census-Income (KDD) Data Set.  Consisted of nearly 300,000 observations of 40 different variables of weighted census data.  Weighted, meaning that each combination of the 40 different variables was given a weight equivalent to the number of persons in the U.S. population with those attributes.

The Census data is from 1994 and 1995, and the following graphs are averaged between the two years.  

Wanted to explore the different effects of race and gender on the representation of population with a full-time income greater than $50,000.

Required some cleaning and filtering to use only the full-time income census observations.

##Income by Sector


```{r, echo = FALSE, message = FALSE, warning = FALSE}
classPlot
```

##Between Males and Females

```{r, echo = FALSE, message = FALSE, warning = FALSE}
classPlot_sex
```

##Questions

-Why is there such a significant difference between males and females among those who make >50k?

-Do we see similar effects between races?

-Keep in mind that all graphs only include those with full-time jobs.

##Income by Race

```{r, echo = FALSE, message = FALSE, warning = FALSE}
racePlot
```

##Income by Race and Sex

```{r, echo = FALSE, message = FALSE, warning = FALSE}
racePlot_sex
```

##Income by age

```{r, echo = FALSE, message = FALSE, warning = FALSE}
age_Plot
```

##More Questions

-Could there be a difference in education?

-Why is there a dip at ~70 years old?

##Population Education Stats

```{r, echo = FALSE, message = FALSE, warning = FALSE}
educationPlot_both
```

##Between Male and Female

```{r, echo = FALSE, message = FALSE, warning = FALSE}
educationPlot_mf
```

