---
title: "Airline Cancellations, Delays, and Geography"
author: "Michael Winfield"
date: "April 24, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## DATASET

The dataset consists of on-time performance statistics from the Bureau of Transportation Statistics ranging across the month of January in 2016.
(http://1.usa.gov/1YP4p5m)

```{r environment, echo = F, message=FALSE, warning=F}
library(ggplot2)
library(dplyr)
library(googleVis)
suppressPackageStartupMessages(library(googleVis))
```
Once I uploaded the dataset, I converted the airline i.d.s to factors, to make using ggplot easier.
```{r code, echo = TRUE, message=FALSE, warning=F }
airdata2 <- read.csv("~/RClass1/airdata2.csv")
airdata2$AIRLINE_ID <- as.factor(airdata2$AIRLINE_ID)
flightdate <- airdata2
```


## Are departure delays related to arrival delays?
```{r delays, echo = F, message=FALSE, warning=F}
flightdate2 <- select(flightdate, FL_DATE, totaldepd = sum(DEP_DELAY_NEW, na.rm=T), totalarrd = sum(ARR_DELAY_NEW, na.rm=T))
ggplot(flightdate2,aes(x=totaldepd, y=totalarrd)) + geom_point()
```

## Cancellations and Flight Date
Did any particular days experience a higher ratio of cancellations per number of flights than others?
```{r plot2, echo = F, message=FALSE, warning=F}
flightdate2 <- group_by(flightdate, FL_DATE)
flight1 <- summarise(flightdate2, totalcancel = sum(CANCELLED), totalflights = sum(FLIGHTS), ratio = totalcancel/totalflights) %>% arrange(., desc(ratio))
ggplot(flight1, aes(x=ratio, y=FL_DATE)) + geom_point()
```

## Blizzard pattern for large airlines
I looked at Southwest Airlines (19393), American Airlines (19805), and Delta Airlines (19790).
```{r plot3, echo = F, message=FALSE, warning=F}
flightdate2 <- group_by(flightdate, FL_DATE, AIRLINE_ID) %>% mutate(., cancel = sum(CANCELLED)) %>% select(., FL_DATE, AIRLINE_ID, cancel)
g <- filter(flightdate2, AIRLINE_ID == '19393' | AIRLINE_ID == '19805' | AIRLINE_ID == '19790') %>% select(., AIRLINE_ID, cancel) %>% group_by(., AIRLINE_ID)
ggplot(g,aes(x=cancel, y=FL_DATE)) + geom_point() + facet_wrap( ~ AIRLINE_ID)
```

## Ratio of cancelled to total flights

The pattern is more severe for the larger airlines than the smaller ones, Alaska Airlines (19930) and Hawaiian Airlines (19690), with Alaska Airlines having a relative spike on the blizzard days.

## Ratio of cancelled to total flights
```{r plot4, echo = F, message=FALSE, warning=F}
flightdate2 <- group_by(flightdate, FL_DATE, AIRLINE_ID) %>% mutate(., cancel = sum(CANCELLED)/sum(FLIGHTS)) %>% select(., FL_DATE, AIRLINE_ID, cancel)
g <- filter(flightdate2, AIRLINE_ID == '19393' | AIRLINE_ID == '19805' | AIRLINE_ID == '19790' | AIRLINE_ID == '19690' | AIRLINE_ID == '19930') %>% select(., AIRLINE_ID, cancel) %>% group_by(., AIRLINE_ID)
ggplot(g,aes(x=cancel, y=FL_DATE)) + geom_point() + facet_wrap( ~ AIRLINE_ID)
```

## Geography and flight patterns
During the blizzard, many larger airline flights originated from blizzard states (Georgia, Florida, North Carolina, Texas).

```{r plot5, echo = F, message=FALSE, warning=F}
blizzard <- filter(flightdate, FL_DATE == '2016-01-23' | FL_DATE == '2016-01-24'|FL_DATE == '2016-01-22' | FL_DATE == '2016-01-25'|FL_DATE == '2016-01-26')
g <- filter(blizzard, AIRLINE_ID == '19393' | AIRLINE_ID == '19805' | AIRLINE_ID == '19790') %>% select(., ORIGIN_STATE_ABR, AIRLINE_ID)
ggplot(data = g, aes(x = ORIGIN_STATE_ABR)) + geom_bar(aes(fill= AIRLINE_ID), position = "stack") 
```

## Smaller airlines
Hawaiian Airlines' and Alaska Airlines' flights fit a different pattern.

```{r plot6, echo = F, message=FALSE, warning=F}
blizzard <- filter(flightdate, FL_DATE == '2016-01-23' | FL_DATE == '2016-01-24'|FL_DATE == '2016-01-22' | FL_DATE == '2016-01-25'|FL_DATE == '2016-01-26')
g <- filter(blizzard, AIRLINE_ID == '19690' | AIRLINE_ID == '19930') %>% select(., ORIGIN_STATE_ABR, AIRLINE_ID)
ggplot(data = g, aes(x = ORIGIN_STATE_ABR)) + geom_bar(aes(fill= AIRLINE_ID), position = "stack")
```

## For all airlines
Where did most of the blizzard cancellations occur?
```{r plot7, echo = F, message=FALSE, warning=F}
blizzard <- filter(flightdate, FL_DATE == '2016-01-23' | FL_DATE == '2016-01-24'|FL_DATE == '2016-01-22' | FL_DATE == '2016-01-25'|FL_DATE == '2016-01-26')
x <- group_by(blizzard, FL_DATE, ORIGIN_STATE_ABR) %>% summarise(., cancel = sum(CANCELLED))
ggplot(x,aes(x=ORIGIN_STATE_ABR, y=cancel)) + geom_point() + coord_polar()
```
Florida, North Carolina, Maryland, New York, and Virginia. 

## Departure delays and geography.1
Do departure delays have the same pattern as cancelled flights?
```{r plot8, echo = F, message=FALSE, warning=F}
blizzard <- filter(flightdate, FL_DATE == '2016-01-23' | FL_DATE == '2016-01-24'|FL_DATE == '2016-01-22' | FL_DATE == '2016-01-25'|FL_DATE == '2016-01-26')
flightdate2 <- group_by(blizzard, ORIGIN_STATE_ABR)
flight1 <- mutate(flightdate2, totaldelay = sum(DEP_DELAY_NEW, na.rm=T) * sum(FLIGHTS)) %>% select(.,ORIGIN_STATE_ABR, totaldelay) %>% summarise(., ddelay = max(totaldelay)) %>% arrange(., desc(ddelay)) 
ggplot(flight1,aes(x=ORIGIN_STATE_ABR, y=ddelay)) + geom_point() + coord_polar()
```

## Departure delays and geography.2
Why does California, a state unaffected by the blizzard, have worse departure delay times than Georgia, Florida, and Texas?


[California1](http://127.0.0.1:10217/custom/googleVis/GeoChartID17c430a64009.html)
```{r plot9, echo = F, message=FALSE, warning=F}
blizzard <- filter(flightdate, FL_DATE == '2016-01-23' | FL_DATE == '2016-01-24'|FL_DATE == '2016-01-22' | FL_DATE == '2016-01-25'|FL_DATE == '2016-01-26')
flightdate2 <- group_by(blizzard, ORIGIN_STATE_ABR)
flight1 <- mutate(flightdate2, totaldelay = sum(DEP_DELAY_NEW, na.rm=T) * sum(FLIGHTS)) %>% select(.,ORIGIN_STATE_ABR, totaldelay) %>% summarise(., ddelay = max(totaldelay)) %>% arrange(., desc(ddelay))
G3 <- gvisGeoChart(flight1, "ORIGIN_STATE_ABR", "ddelay", options=list(region="US", displayMode="regions", resolution="provinces", width=600, height=400))
plot(G3)
```

## A California-local phenomenon?

I suspect that it is because of where the California flights are headed, so I looked to the arrival delays for flights headed from California during the blizzard.


[California2](http://127.0.0.1:10217/custom/googleVis/GeoChartID17c415ec4c24.html)
```{r plot10, echo = F, message=FALSE, warning=F}
blizzard <- filter(flightdate, FL_DATE == '2016-01-23' | FL_DATE == '2016-01-24'|FL_DATE == '2016-01-22' | FL_DATE == '2016-01-25'|FL_DATE == '2016-01-26')
CAdep <- filter(blizzard, ORIGIN_STATE_ABR == 'CA') %>% group_by(., DEST_STATE_ABR) %>% mutate(., totalarrd = sum(ARR_DELAY_NEW, na.rm=T)) %>% select(., DEST_STATE_ABR, totalarrd) %>% summarise(., arrdelay = max(totalarrd))
G3 <- gvisGeoChart(CAdep, "DEST_STATE_ABR", "arrdelay", options=list(region="US", displayMode="regions", resolution="provinces", width=600, height=400))
plot(G3)
```
The highest proportion of flights leaving California during the blizzard that landed late did so in California.