---
title: "Research and development in various countries"
author: "Yannick Kimmel"
date: "April 24, 2016"
output: 
ioslides_presentation:
    widescreen: true
    incremental: true
    css: assets/css/ioslides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

* Research & development (R&D) can increase the productivity of a nation. 
* Personally became aware of the lack of R&D jobs available in manufacturing when finishing PhD.
* The amount of PhD graduates competing for professorships.  
* Want to understand how this is affecting the country.
* Understanding how this is changing is important to policy makers, economists, and researchers. 
```{r set, echo=FALSE}
setwd("/Users/YannickMac/Dropbox/Applications/Data science/NYCDSA/Data_visualization_project/Yannick Kimmel")
library(dplyr)
library(ggplot2)
indicators = read.csv("subbeddata.csv", header = TRUE, stringsAsFactors = FALSE)
countries = c("EUU", "USA", "CHN", "JPN", 
              "RUS", "IND", "KOR") #WLD, OED

indicatorcodes = c("GB.XPD.RSDV.GD.ZS", "SP.POP.SCIE.RD.P6", 
                   "IP.JRN.ARTC.SC", "IP.PAT.NRES", "IP.PAT.RESD", "IP.TMK.TOTL",
                   "NV.IND.MANF.ZS")
subdata = function(indicatorCode) {
    filter(indicators, IndicatorCode == indicatorCode) %>%
        filter(CountryCode %in% countries)
}
```
## World development indicators dataset from World Bank 

* Statistics on up to 247 countries and regions on 1,344 development indicators 
* Yearly indicators as early as 1960 to 2014
* Countries of interest : China, Euro Union, India, Japan, Korea, Russia, USA
*Easy to use 

example =  filter(indicators, IndicatorCode == indicatorCode) %>%
        filter(CountryCode %in% countries)


## Input: R&D expenditure as % of GDP
```{r R&D expenditure}
researchexpn =  subdata("GB.XPD.RSDV.GD.ZS")
res <- ggplot(researchexpn, aes(x = Year, y = Value)) +geom_line(aes(color = CountryCode), size = 1)
res + theme_bw() + ggtitle("R&D expenditures as a percentage of GDP") +
    scale_color_discrete(labels = c("China","Euro Union", "India",
                                    "Japan", "Korea", "Russia", "USA")) + 
    guides(color=guide_legend(title="Countries")) 
```
* Korea and China rapidly increasing. 
* Japan and USA steady and high. 

## R&D output: Journal publications
* Mostly an academic measure. 
* Measure of quantity not quality. 

```{r publications}
publications = subdata("IP.JRN.ARTC.SC")
pub = ggplot(publications, aes(x = Year, y = Value)) +geom_line(aes(color = CountryCode), size = 1)
pub + theme_bw() +  ggtitle("Number of scientific and technical journal articles") +
    scale_color_discrete(labels = c("China","Euro Union", "India",
                                    "Japan", "Korea", "Russia", "USA")) + 
    guides(color=guide_legend(title="Countries")) 
```
* Euro Union and USA publish a high amount. Euro Union increasing. 
* Japan on decline. 
* China rapidly rising.

## R&D output: Patent applications
* Mostly an industrial measure. 
* Measure of quantity not quality. 
```{r patents}
patents = filter(indicators, IndicatorCode == "IP.PAT.NRES" | 
                     IndicatorCode == "IP.PAT.RESD") %>%
    filter(CountryCode %in% countries) %>% group_by(CountryCode, Year) %>%
    summarise(totpap = sum(Value))
pat = ggplot(patents, aes(x = Year, y = totpap)) + geom_line(aes(color = CountryCode), size = 1)
pat + theme_bw() + ggtitle("Number of patents") + ylab("total patents") +xlim(1980, 2014) +
    scale_color_discrete(labels = c("China","Euro Union", "India",
                                    "Japan", "Korea", "Russia", "USA")) + 
    guides(color=guide_legend(title="Countries")) 
```
* China increasing rapidly. USA also high. 

## Output: Trademark applications
* Software mostly not protected by patents
```{r trademarks}
trademark = subdata("IP.TMK.TOTL")
trade = ggplot(trademark, aes(x = Year, y = Value)) + geom_line(aes(color = CountryCode), size = 1)
trade + theme_bw() + ggtitle("Number of trademark applications") +
    scale_color_discrete(labels = c("China","Euro Union", "India",
                                    "Japan", "Korea", "Russia", "USA")) + 
    guides(color=guide_legend(title="Countries")) 
```
* China rapidly increasing. USA also increasing.

## Number of Researchers in R&D 

```{r researchers}
researchers = subdata("SP.POP.SCIE.RD.P6")
respeople <- ggplot(researchers, aes(x = Year, y = Value)) +geom_line(aes(color = CountryCode), size = 1)
respeople + theme_bw() + ggtitle("Number of Researchers in R&D per million people") +
    scale_color_discrete(labels = c("China","Euro Union", "India",
                                    "Japan", "Korea", "Russia", "USA")) + 
    guides(color=guide_legend(title="Countries")) 
```

## Value added manufacturing

```{r manufacturing}
manufacturing = subdata("NV.IND.MANF.ZS")
manufact = ggplot(manufacturing, aes(x = Year, y = Value)) + geom_line(aes(color = CountryCode), size = 1)
manufact + theme_bw() + ggtitle("Manufacturing, value added (% of GDP)") +
    scale_color_discrete(labels = c("China","Euro Union", "India",
                                    "Japan", "Korea", "Russia", "USA")) + 
    guides(color=guide_legend(title="Countries")) 
```

##Conclusions
* China’s R&D expenditure and output rapidly increasing. 
* USA and EU are increasing at a lower rate.

* Future work:
- Correlating economic performance with R&D. 