---
title: "Restaurant health inspections in NYC"
author: "Ho Fai Wong"
date: "April 24, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#Load libraries
library(plyr)
library(dplyr)
library(zipcode)
library(ggplot2)
library(choroplethrZip)
library(lubridate)

#Load inspection data
raw = read.csv("./DOHMH_New_York_City_Restaurant_Inspection_Results.csv", stringsAsFactors = F)
raw.df = tbl_df(raw)

#Rename and re-format columns
names(raw.df) = tolower(names(raw.df))
raw.df = rename(raw.df, cuisine = cuisine.description) %>%
  mutate(., inspection.date = as.Date(inspection.date, "%m/%d/%Y")) %>%
  mutate(., grade.date = as.Date(grade.date, "%m/%d/%Y")) %>%
  mutate(., record.date = as.Date(record.date, "%m/%d/%Y")) %>%
  mutate(., phone = as.double(phone)) %>%
  mutate(., boro = factor(boro, levels(factor(boro))[c(1:3,5:6,4)])) %>%
  mutate(., zipcode = clean.zipcodes(zipcode)) %>% #Turn zip to string with clean format e.g. preceding 0s, for zip_choroplethr

  #Shorten/reword action and cuisine
  mutate(., cuisine = gsub(pattern = 'Latin \\(Cuban, Dominican, Puerto Rican, South \\& Central American\\)', replacement = 'Latin', x = cuisine, ignore.case = F)) %>%
  mutate(., cuisine = gsub(pattern = 'Café/Coffee/Tea', replacement = 'Cafe/Coffee/Tea', x = cuisine, ignore.case = F)) %>%
  mutate(., cuisine = factor(cuisine, levels(factor(cuisine))[c(1:55,57,59:84,58,56)])) %>% #Moving N/A and Other to the bottom of cuisine factors
  mutate(., action = gsub(pattern = "Violations were cited in the following area\\(s\\).", replacement = "violations", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "No violations were recorded at the time of this inspection.", replacement = "no violations", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "Establishment re-opened by DOHMH", replacement = "reopened", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "Establishment Closed by DOHMH.  violationsand those requiring immediate action were addressed.", replacement = "closed", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "Establishment re-closed by DOHMH", replacement = "reclosed", x = action, ignore.case = F)) %>%

  #Add columns for analysis
  mutate(., yearmon = as.Date(paste("1",
                                    strftime(inspection.date, "%m"), 
                                    strftime(inspection.date, "%Y"), 
                                    sep="."), 
                              format="%d.%m.%Y")) %>% #Add column for month and year of inspection
  mutate(., new_grade = ifelse(score < 0, 'Negative', 
                               ifelse(score < 14 , 'A',
                                      ifelse(score < 28, 'B', 'C')))) #Assign grades based on scores (existing data may have score but no grade)

  #Fix bad data
levels(raw.df$boro) = c('Bronx','Brooklyn','Manhattan','Queens','Staten Island','Missing')
raw.df$boro[raw.df$zipcode==11249] = 'Brooklyn' #Fill missing data...
raw.df = filter(raw.df, 
                  zipcode!='07005' & #Eliminate NJ inspection...
                  boro != 'Missing' & #Eliminate remaining rows with boro == 'Missing'; confirmed that none are in NYC
                  inspection.date > '1900-01-01' & #Eliminate rows where the restaurant hasn't been inspected yet
                  !is.na(score) & #Eliminate rows without a score
                  score >= 0 #Eliminate rows with a negative score
)


#Unique inspections
inspections = unique(select(raw.df, camis, boro, zipcode, cuisine, 
                            inspection.date, action, score, new_grade, yearmon))

#Unique inspections but only the latest for each restaurant
latest = merge(
  aggregate(inspection.date ~ camis, inspections, max), 
  inspections)

#Unique restaurants and their latest score/grade
restaurants = unique(select(latest, camis, boro, zipcode, cuisine, score, new_grade))
```

## Introduction

* NYC Health Department restaurant **inspections, scores and grades** (e.g. A/B/C)

* How do a restaurant's **location** and type of **cuisine** affect its inspection results? E.g.:
    + Do restaurants in Manhattan have better scores than Queens or the Bronx?
    + Do restaurants in Chinatown and Flushing perform worse than Upper East Side?
    + Do Chinese restaurants have poorer hygiene than French or American restaurants?

* Data used (after cleanup and filtering): 2010 - 2016; 150k inspections (rows); 25k restaurants

## Grades by borough

```{r}
#Bar plot: Restaurants by borough and latest grade
ggplot(data=latest, aes(x=reorder(boro, boro, function(x)-length(x)))) + 
  geom_bar(aes(fill=new_grade), position='dodge') +
  labs(title='Restaurants by borough and latest grade', 
       x='Borough', 
       y='Restaurants') +
  scale_y_continuous(limits=c(0,9000), breaks=seq(0,10000,1000)) +
  scale_fill_brewer(name="Grade", palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```

## Going deeper... Scores by borough

```{r}
#Density: Restaurants by latest score and borough
ggplot(data=latest, aes (x=score)) + 
  stat_density(aes(color=boro), geom="line",position="identity") + 
  coord_cartesian(xlim=c(0,40)) +
  labs(title='Density of restaurants by latest score and borough', 
       x='Score', 
       y='Restaurant density') + 
  scale_y_continuous(breaks=seq(0,0.14,0.02),
                     labels = scales::percent) +
  scale_colour_brewer(name="Borough", palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank()) +
  geom_vline(xintercept = c(14,28), colour='grey') +
  annotate("text", x = c(6,20,35), y = 0.11, label = c('A','B','C'), size=6) +
  annotate("rect", xmin = 0, xmax = 14, ymin = 0, ymax = 0.13, alpha = .2, fill='darkred') +
  annotate("rect", xmin = 14, xmax = 28, ymin = 0, ymax = 0.13, alpha = .2, fill='darkblue') +
  annotate("rect", xmin = 28, xmax = 60, ymin = 0, ymax = 0.13, alpha = .2, fill='darkgreen')
```

## Going deeper still... Scores by zipcode

```{r, warning = FALSE, message = FALSE, fig.height=6}
#Map average scores by zipcode
avgbyzip = latest %>% 
  group_by(., zipcode) %>%
  summarise(., value=mean(score)) %>%
  rename(., region=zipcode)

#Choroplethr doesn't have all zipcodes in the NY inspection data (uses 2010 census ZCTAs)
scored_zips = unique(avgbyzip$region) #zipcodes that appear in inspection data data
missing_zips = c('10048','10055', '10057', '10104', '10105', '10106', '10107', '10118', '10121', '10123', '10155', '10166', '10175', '10176', '10178', '10179', '10281', '10285', '10317', '11242', '11249', '11256', '11352') #zipcodes that are not in choropleth package
adjusted_zips = scored_zips[!(scored_zips %in% missing_zips)] #zipcodes that appear in original data and in choropleth package

#Map: Average scores by zipcode; zipcodes are binned into 9 buckets of equal size due to limitations of the function
zip_choropleth(avgbyzip, 
               zip_zoom = adjusted_zips,
               title="Average of latest scores by zipcode",
               num_colors=5) + 
  scale_fill_brewer(palette='OrRd', name='Average score')
```


## What about inspection closures?

* Scores don't tie directly to restaurant closures (e.g. public health hazard)

* Defined the following ratios:
    + **Inspection closure ratio**: % of inspections that lead to the restaurant being closed
    + **Repeat closure ratio**: % of restaurants that were closed during more than one inspection cycle

## Closures by borough

```{r}
#Total counts of inspections that led to closures by borough
inspClosByBoro = inspections %>%
  group_by(., boro) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

#Restaurants that were closed at least once (closures, not reclosures as part of the same inspection cycle)
reclosedByBoro =  inspections %>% 
  filter(., action == 'closed') %>%
  group_by(., camis, boro, zipcode) %>%
  summarize(., count = n()) %>%
  group_by(., boro) %>%
  summarize(., total_closed = n(), 
            closed_more_than_once = sum(count>1), 
            ratio = closed_more_than_once / total_closed)

#Closure and repeat closure combined - by borough
boroClosureRatios = left_join(inspClosByBoro, reclosedByBoro, by = 'boro') %>%
  rename(., inspRatio = ratio.x, recloRatio = ratio.y)

ggplot(data = boroClosureRatios, aes(x=inspRatio, y=recloRatio)) + 
  geom_point(aes(colour = boro), size = 3) +
  labs(title='Inspection closure and repeat closure ratios by borough',
       x='Inspection closure ratio',
       y='Repeat closure ratio') +
  scale_x_continuous(limits = c(0.01,0.025),
                     labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_brewer(palette = 'Set1') + 
  theme_bw() +
  theme(legend.position='none') +
  geom_text(aes(label=boro), 
            check_overlap=TRUE,
            size=3,
            hjust = 0, 
            nudge_x = 0.0003)
```


## Closures by cuisine

```{r, warning = FALSE}
#Down-select top 20 cuisine types for simpler visualization (out of 84)
restByCuisineFiltered = restaurants %>%
  group_by(., cuisine) %>%
  summarise(., count=n()) %>%
  arrange(., desc(count)) %>%
  top_n(., 20, count)

restaurantsFiltered = tbl_df(semi_join(restaurants, restByCuisineFiltered, by = 'cuisine')) #Keep top 20 covering ~81%
inspectionsFiltered = tbl_df(semi_join(inspections, restByCuisineFiltered, by = 'cuisine'))

#Ratio of inspection closures by top 20 cuisine types
inspClosByCuisine = inspectionsFiltered %>%
  group_by(., cuisine) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

#Restaurants that were closed at least once (closures, not reclosures as part of the same inspection cycle)
reclosedByCuisine = unique(
  inspectionsFiltered %>%
    filter(., action == 'closed') %>%
    select(., camis, cuisine, inspection.date)) %>%
  group_by(., camis, cuisine) %>%
  summarise(., closures = n()) %>% 
  group_by(., cuisine) %>%
  summarise(., closed_more_than_once = sum(closures>1), 
            total_closed = n(),
            ratio = closed_more_than_once / total_closed) 

#Closure and repeat closure combined - by cuisine
cuisineClosureRatios = left_join(inspClosByCuisine, reclosedByCuisine, by = 'cuisine') %>%
  rename(., inspRatio = ratio.x, recloRatio = ratio.y) %>%
  left_join(., (restaurants[,c('cuisine','camis','score')] %>% 
                  group_by(., cuisine) %>% 
                  summarise(., count = n(), avg = mean(score))),
            by = 'cuisine') #Added restaurant count by cuisine
```


```{r, eval=FALSE}
ggplot(data = cuisineClosureRatios[cuisineClosureRatios$recloRatio>=0,], aes(x=inspRatio, y=recloRatio)) + 
  geom_point(aes(colour=avg), size = 2) +
  labs(title='Inspection closure and repeat closure ratios by top 20 cuisine',
       x='Inspection closure ratio',
       y='Repeat closure ratio') +
  scale_x_continuous(limits=c(0,0.03),
                     labels = scales::percent) +
  scale_y_continuous(limits=c(0,0.2),
                     labels = scales::percent) +
  scale_colour_gradient(name='Average score',
                        low='yellow',
                        high='red') +
  theme_bw() +
  theme(legend.key=element_blank(), legend.position='bottom') +
  geom_text(aes(label=cuisine),
            size=5,
            hjust = 0, 
            nudge_x = 0.0002,
            nudge_y = 0.003,
            angle=45)
```

```{r fig.width=8, fig.height=5, echo=FALSE}
#Inserting exported image of the previous plot for better visibility
library(png)
library(grid)
img <- readPNG("./closureRatioByCuisine.png")
 grid.raster(img)
```

## Closures by cuisine and borough

```{r}

#Ratio of inspection closures by top 20 cuisine and boro
inspClosByCuisineBoro = inspectionsFiltered %>%
  group_by(., cuisine, boro) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

#Refactor cuisines to order by descending count of restaurants by cuisine
restFiltCount = restaurantsFiltered %>%
  group_by(., cuisine) %>%
  summarise(., count = n())
inspClosByCuisineBoro$cuisine = factor(inspClosByCuisineBoro$cuisine)
newFactor = levels(inspClosByCuisineBoro$cuisine)[order(-rank(restFiltCount$count))]
inspClosByCuisineBoro$cuisine = factor(inspClosByCuisineBoro$cuisine, levels=rev(newFactor))

#Faceted bar plot: Inspection closure ratio by top 20 cuisine and borough
ggplot(data=inspClosByCuisineBoro, aes(x=cuisine, y=ratio)) + 
  geom_bar(stat='identity', aes(fill=boro)) + 
  facet_grid(. ~ boro) +
  labs(title='Inspection closure ratio by top 20 cuisine and borough', x='Top 20 cuisine', y='Inspection closure ratio') +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = 'Set1', name = 'Borough') +
  theme_bw() +
  theme(legend.key=element_blank(), legend.position="bottom") +
  coord_flip()

```


## Conclusion

* Displaying scores in addition to grades could further improve hygiene (most have A already)
* My specific initial questions were addressed...
    + Scores differentiate neighborhoods but not boroughs; closure ratios provide additional comparisons
    + Chinatown is on par with UES but Flushing is not
    + Chinese restaurants have higher closure ratios than French and American restaurants
* ... but more observations can be made based on your culinary and geographic selections, e.g:
    + Brooklyn and the Bronx have the worst rates of inspection and repeat closures respectively
    + Asian restaurants in the Bronx and Staten Island have a much higher inspection closure ratio, etc
    
## Ideas for further analysis

* Investigate violation types
* Compare with NYC demographic data by neighborhood
* Analyze trends over time by neighborhood and cuisine
* Correlate with popularity of restaurants

# Appendix

## Scores by borough

```{r}
#Box plot: Latest score by borough
ggplot(data=latest, aes(x=reorder(boro, boro, function(x)-length(x)), y=score)) +
  geom_boxplot(aes(fill=boro)) +
  labs(title='Latest score by borough',
       x='Borough',
       y='Latest score') +
  stat_summary(fun.y=mean, colour="white", geom="point", shape=16, size=2) +
  scale_fill_brewer(name="Borough", palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```


## Restaurants by score and borough

```{r}
#Freqpoly: Restaurants by latest score and borough
ggplot(data=latest, aes(x=score)) +
  geom_freqpoly(aes(color=boro), binwidth=2) +
  coord_cartesian(xlim=c(0,40)) +
  labs(title='Restaurants by latest score and borough',
       x='Latest score',
       y='Restaurants') +
  scale_colour_brewer(name="Borough", palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```


## Inspection closure scores by borough

```{r}
#Score of restaurants closed by inspections
ggplot(data = inspections[inspections$action %in% c('closed','reclosed'),], aes(x=boro, y=score)) +
  geom_boxplot(aes(fill=boro)) +
  labs(title='Score of restaurants closed by inspections by borough',
       x='Borough',
       y='Latest score') +
  stat_summary(fun.y=mean, colour="white", geom="point", shape=16, size=2) +
  scale_fill_brewer(name="Borough", palette='Set1') + 
  theme_bw() +
  theme(legend.key=element_blank())
```


## Average scores by cuisine (1/2)

```{r}
#Restaurants by average score and cuisine
avgScoreByCuisine = latest %>%
  group_by(., cuisine) %>%
  summarise(., avg = mean(score)) %>%
  arrange(., desc(avg))

#Bar plot: Average score by cuisine
ggplot(data=avgScoreByCuisine[1:42,], aes(x=reorder(cuisine, avg), y=avg)) +
  geom_bar(stat='identity') +
  labs(title='Average score by cuisine',
       x='Cuisine',
       y='Average score') +
  theme_bw() +
  coord_flip()
```


## Average scores by cuisine (2/2)

```{r}
ggplot(data=avgScoreByCuisine[43:84,], aes(x=reorder(cuisine, avg), y=avg)) +
  geom_bar(stat='identity') +
  labs(title='Average score by cuisine',
       x='Cuisine',
       y='Average score') +
  theme_bw() +
  coord_flip()
```


## Restaurants by cuisine and grade

```{r}
#Stacked bar plot: Restaurants by cuisine and grade
ggplot(data=restaurantsFiltered, aes(x=reorder(cuisine, cuisine, function(x) length(x)))) +
  geom_bar(aes(fill=new_grade)) +
  coord_flip() +
  labs(title='Restaurants by top 20 cuisine and grade',
       x='Top 20 cuisine',
       y='Restaurants') +
  scale_fill_discrete(name='Grade') +
  theme_bw() +
  theme(legend.key=element_blank())
```


## Restaurants by score and cuisine

```{r}
#Density: Restaurants by score and top 20 cuisine
ggplot(data=restaurantsFiltered, aes (x=score)) +
  stat_density(aes(color=cuisine), geom="line",position="identity") + 
  coord_cartesian(xlim=c(0,40)) +
  labs(title='Density of restaurants by score and top 20 cuisine', 
       x='Score', 
       y='Density') +
  scale_colour_discrete(name="Cuisine") +
  scale_y_continuous(limits=c(0,0.15), labels = scales::percent) +
  theme_bw() +
  theme(legend.key=element_blank(),
        legend.key.size=unit(0.4,"cm"))

```


## Trend of score by borough

```{r}
trendScoreYear = inspections %>%
  mutate(., year = as.numeric(format(inspection.date, '%Y'))) %>%
  group_by(., year, boro) %>%
  summarise(., avg = mean(score))

ggplot(data=trendScoreYear, aes(x=year, y=avg)) + 
  geom_freqpoly(stat='identity', aes(colour=boro)) +
  labs(title='Average score by borough over time',
       x='Year',
       y='Average score') +
  coord_cartesian(ylim = c(4,20)) +
  scale_x_continuous(breaks = 2010:2016) +
  scale_color_brewer(name='Borough', palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```


## Trend of score by month

```{r}
#Trend of average scores by month and borough
trendScoreMonth = inspections %>%
  mutate(., month=month(inspection.date)) %>%
  group_by(., month, boro) %>%
  summarise(., avg = mean(score))

ggplot(data=trendScoreMonth, aes(x = month, y = avg)) + 
  geom_freqpoly(stat='identity', aes(color=boro)) +
  labs(title='Average score by month and borough',
       x='Month',
       y='Average score') +
  coord_cartesian(xlim = c(1,12), ylim = c(13,18)) +
  scale_x_continuous(breaks = 1:12, 
                     labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")) +
  scale_color_brewer(name='Borough', palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())

```


## Trend of inspection closure by month

```{r}
#Trend of inspection closure ratio by month and borough
trendClosures = inspections %>% 
  mutate(., month=month(inspection.date)) %>%
  group_by(., month, boro) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

ggplot(data=trendClosures, aes(x=month, y=ratio)) + 
  geom_freqpoly(stat='identity', aes(color=boro)) + 
  labs(title='Inspection closure ratio by month and borough',
       x='Month',
       y='Inspection closure ratio') +
  coord_cartesian(xlim = c(1,12)) +
  scale_x_continuous(breaks = 1:12, 
                     labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")) +
  scale_y_continuous(limits=c(0.005,0.035),
                     breaks=seq(0,0.035,0.005),
                     labels=scales::percent) +
  scale_color_brewer(name='Borough', palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```

