---
title: "Visualizing the Impact of Palm Oil Production"
author: "Wanda Wang"
date: "April 24, 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction
Deforestation is accelerating at an alarming rate around the world as a result of Palm oil production. In order to satisfy growing global demand for Palm oil, an ingredient commonly found in cosmetics to processed foods - clearing the way for lucrative oil plantations by forest fires is negatively impacting our environment. Concerns include the loss of rainforest habitat for endangered species, harmful carbon dioxide emissions, coupled with the displacement of local indigenous communities. 

##Exploratory Questions
What countries are the top producers of Palm oil? 
Where are the palm oil mills located? 
Where are the forest-fire hotspots? 

##The Datasets
The UN Food and Agriculture organization provides information on global Palm oil production and area harvested. The Global Forest Watch also has Palm oil Mill location data and fire alert data available for our review.

##Industry Overview
To navigate to where Palm Oil production is at its highest volume, we first approach the UN Food and Agriculture dataset: 
```{r}
setwd("/Users/drscholls303/Desktop/Project1")
# Load the required R packages and specified datasets
library(dplyr)
library(ggplot2)
faostat.palm <-read.csv("Faostats.csv")
faostat.palm <- faostat.palm %>% na.omit()
#Select relevant columns
producers <-select(faostat.palm, AreaName, ElementName, Value, Year)
# Filter for top producers in most recent available year of 2014
producers.year<-filter(producers, Year == 2014) %>% 
  group_by(AreaName, ElementName='Production') %>% 
  summarise(., Production=sum(Value)) 
nrow(producers.year) #46 rows 
#Calculate percentage of world production for each country
producers.year$Percent = round(producers.year$Production[1:46]*100/
          sum(producers.year$Production), digits = 2)
producers.year<-tbl_df(producers.year)
top<-top_n(producers.year,5,Production)
top
#top 5 graph
library(scales)
top.plot<-ggplot(top, aes(AreaName, Production, fill=Percent)) + geom_bar(stat="identity") + xlab("Country") + ylab("Oil Production") +
  ggtitle("Palm Oil Production by Country (Top 5)")
  
firstplot<- top.plot + 
scale_y_log10(
breaks = scales::trans_breaks("log10", function(x) 10^x),
labels = scales::trans_format("log10", math_format(10^.x)))
firstplot
```
The top five producers in the world include Indonesia and Malaysia. Indonesia accounts for over 46% or worldwide production. Malaysia accounts for around 34%. The combined 80% allocation in this region should be noted. 

## Industry Map - Bird's eye view 
```{r}
library(maptools)
library(ggplot2)
library(ggmap)
library(rgeos)
library(dplyr)
plotmap <- readShapePoly(fn="TM_WORLD_BORDERS-0.3/TM_WORLD_BORDERS-0.3.shp")
palmHa <-read.csv("Faostats.csv", stringsAsFactors=F)
palmHa <-filter(palmHa, ElementName == 'Area harvested')
names(plotmap)[5] <- "AreaName"
target <- c("Indonesia", "Malaysia", "Colombia", "Thailand", 
            "Philippines", "Nigeria", "Cameroon", "Ecuador", "Ghana", "Honduras","Guatemala")
#Use foritfy to transform spatial plotmap into dataframe
plotmapDf <- fortify(plotmap, region = "AreaName") %>% 
  filter(id %in% target)
palmHaMapDf <- merge(plotmapDf, palmHa, by.x="id", by.y="AreaName")
backgroundMap <- ggplot(data=palmHaMapDf) + geom_path(aes(x=long, y=lat, group=group), color='grey') + coord_equal() + geom_polygon(aes(x=long, y=lat, group=group, fill=Value))
mapAsia <- get_map(location = 'Indonesia', zoom=4)
mapAfrica <- get_map(location = 'Africa', zoom=4)
mapSouthAm <- get_map(location = 'South America', zoom=4)

ggmapObjAsia <- ggmap(mapAsia)
Asia <- ggmapObjAsia + geom_polygon(aes(x=long, y=lat, group=group, fill=Value), data=palmHaMapDf, alpha=.9) + 
  geom_path(aes(x=long, y=lat, group=group), data=palmHaMapDf, color='black')
Asia

ggmapObjAf <- ggmap(mapAfrica)
Africa <- ggmapObjAf + geom_polygon(aes(x=long, y=lat, group=group, fill=Value), data=palmHaMapDf, alpha=.9) + 
  geom_path(aes(x=long, y=lat, group=group), data=palmHaMapDf, color='black')
Africa

ggmapSouthAm <- ggmap(mapSouthAm)
SouthAm <- ggmapSouthAm + geom_polygon(aes(x=long, y=lat, group=group, fill=Value), data=palmHaMapDf, alpha=.9) + 
  geom_path(aes(x=long, y=lat, group=group), data=palmHaMapDf, color='black')
SouthAm
```


As a visual confirmation on the map - looking at SouthEast Asia vs Africa or South America, there are more acres harvested within Indonesia. 


## Palm Oil Harvested in Indonesia

```{r}
library(dplyr)
library(ggplot2)
faostat.palm <-read.csv("Faostats.csv")
faostat.palm <- faostat.palm %>% na.omit()
AH<-filter(faostat.palm, ElementName=='Area harvested', AreaName=='Indonesia')

areaGraph<-ggplot(AH, aes(Year, Value, fill=AreaName)) + geom_bar(stat="identity") +
  ylab("Area Harvested") + ggtitle("Area Harvested through Time(Indonesia)")
areaGraph
```


The rise in Area Harvested over time is alarming over time. The drastic percent change from 2000 to 2014 was over 186%. From a value of 2,014,000 from 740,709. 

## Mill Distribution and Fire Hotspots
```{r}
library(dplyr)
library(leaflet)
mills<-read.csv("mills.csv")
fires<-read.csv("fires.csv")
last.year.fires <-filter(fires, OBJECTID > 11556)#fire-alerts, as of 1/01/2015
#zoom in on SE asia, Circles for mills, Markers for fire alerts
map <- leaflet(mills) %>%
  addTiles('http://{s}.basemaps.cartocdn.com/dark_all/{z}/{x}/{y}.png',attribution='Map tiles by <a href="http://stamen.com">Stamen Design</a>, <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a> &mdash; Map data &copy; <a href="http://www.openstreetmap.org/copyright">OpenStreetMap</a>') %>% addCircles(~longitude, ~latitude, popup=mills$type, weight = 3, radius=40, color="#ffa500", stroke = TRUE, fillOpacity = 0.8) %>%
  addMarkers(data = last.year.fires, ~Latitude, ~Longitude, clusterOptions = markerClusterOptions()) %>% 
  addLegend("bottomright", colors= "#ffa500", labels="Mills", title="Legend") 
map
```

##Conclusion
There is a strong presence of fire hotspots within the last year where the majority of Mills are located. Indonesia possesses the most Palm Oil Mills as currently reported in available data. There are hardly any visible fire hotspots in non-Palm Oil producing countries. 

##Future Questions
Are there any geographic overlaps with existing nature reservations- what is the tree cover loss over time in those areas?
How much has the Air Quality Index been impacted?
Is there a trend between species extinction and habitat location in Palm Oil producing nations? 
 